---
title: "DS PO2PLS two steps analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load data 
## methylation data are after quality control following Bacalini 2015 paper. Cell count is corrected. Age, sex not corrected. Column centered.
## Glycomics data are log TA normalized. Age, sex not corrected. Column centered.
## Samples are overlaped
```{r load data, message=FALSE, results=FALSE, warning=FALSE}
library(OmicsPLS)
library(magrittr)
library(ggplot2)
library(ggforce)
library(PO2PLS)
library(dplyr)
library(lme4)
library(lmerTest)
library(gridExtra)
library(survival)
# source("/home/z/Mygit/Supervised_PO2PLS/SuPO2PLS.R")
source('/home/z/Mygit/Scripts/DOWN/plot_functions.R')
load("~/Data/DOWN/Pre-integrate_count_adj.RData")
gly %<>% scale(scale = F)
# load("~/Data/DOWN/Pre-integrate_DSP_DSS_adj.RData")
probe <- data.table::fread(file = '~/Data/DOWN/GSE52588_family.soft', header = T, skip=215, verbose = F, na.strings="") %>% arrange(ID) # for map CpG to genes

sample_info <- read.delim("/home/z/Data/DOWN/methy-glycan_name.txt", sep = " ", header = T)
sample_info$ID %<>%  as.character()
sample_info %<>% filter(ID %in% rownames(gly))
all.equal(sample_info$ID, rownames(gly))
```

# PCA
```{r, fig.height=5, fig.width=10}
pca_meth <- prcomp(meth, rank. = 56)

pca_plot_meth <- data.frame(pca_meth$x, group=sample_info$group)

# PC 1,2, and 2,3
grid.arrange(grobs = list(pca_plot(pca_plot_meth, xPC=1, yPC=2, sample_info$group, 'PCA: Methylation', pca_meth$sdev), pca_plot(pca_plot_meth, xPC=2, yPC=3, sample_info$group, 'PCA: Methylation', pca_meth$sdev), pca_plot(pca_plot_meth, xPC=3, yPC=4, sample_info$group, 'PCA: Methylation', pca_meth$sdev)), nrow = 1, ncol=3)


pca_gly <- prcomp(gly, rank. = 10)
pca_plot_gly <- data.frame(pca_gly$x, group=sample_info$group)

# PC 1,2, and 2,3
grid.arrange(grobs = list(pca_plot(pca_plot_gly, xPC=1, yPC=2, sample_info$group, 'PCA: Glycomics', pca_gly$sdev), pca_plot(pca_plot_gly, xPC=2, yPC=3, sample_info$group, 'PCA: Glycomics', pca_gly$sdev), pca_plot(pca_plot_gly, xPC=3, yPC=4, sample_info$group, 'PCA: Glycomics', pca_gly$sdev)), nrow = 1, ncol=3)

```


#############################################################
# Now, analysis with all the samples (DS vs controls)

## Nr of components for PO2PLS from scree plot
r,rx,ry: 3,1,1
```{r, fig.height=5, fig.width=10, message=FALSE}
cv <- crossval_o2m_adjR2(meth, gly, 1:3,0:2,0:2, nr_folds = 5);cv
fit <- o2m(meth, gly, 10, 0,0, p_thresh = 1, stripped = T)
D <- (t(fit$W.) %*% t(meth)) %*% (gly %*% fit$C.)
d <- diag(D)#[-1]
#plot(cumsum(d)/sum(d))

d1 <- svd(meth%*%t(meth), 0, 0)$d#[-1]
#plot(cumsum(d2)/sum(d2))

d2 <- svd(gly%*%t(gly), 0, 0)$d#[-1]
#plot(cumsum(d2)/sum(d2))

par(mfrow=c(1,3))
plot(d/sum(d), main='Joint Scree plot')
plot(d1/sum(d1), main="Methylation Scree plot") 
plot(d2/sum(d2), main="Glycomics Scree plot")
par(mfrow=c(1,1))
```


## PO2PLS
```{r, message=FALSE}
set.seed(202101)
# The log likelihood upon convergence is on the scale of 10^7, tol set to 10^-3 to save time.
fit_po2 <- PO2PLS(meth,gly,3,1,1,steps = 1000, tol = 0.001, init_param = 'o2m') 

sd_B <- variances_inner.po2m(fit_po2, meth, gly)
z_B <- fit_po2$parameters$B %>% diag / sd_B
p_B <- 2*(1-pnorm(z_B))
# p-value of B
p_B

# Get joint scores
Tt <- with(fit_po2$parameters, (meth - meth%*%Wo%*%t(Wo)) %*% W)
# Tt_test <- with(fit_po2$params, meth%*%W)
U <- with(fit_po2$parameters, (gly - gly%*%Co%*%t(Co)) %*% C)

jointPC <- data.frame(Tt = Tt, U = U, sample_info %>% select(group, age,sex,family, group_ds))
jointPC$group %<>%  factor(levels = c('DS', 'SB', 'MA'))
```

## distribution of joint PC
```{r,fig.height=7, fig.width=10}
# Density plot
p <- list()
p <- lapply(1:6, function(e) ggplot(jointPC, aes(x=jointPC[,e], color = group)) + geom_density(alpha = 0.5) +
         ggtitle(colnames(jointPC)[e]))

gridExtra::grid.arrange(grobs=p, nrow=2, ncol=3)

# Dot plot on groups
p1 <- ggplot(jointPC, aes(x=group, y=-Tt.1, color = group)) + 
  geom_sina(aes(group = group), show.legend = FALSE) +
  ylab("Methylation (t1)")+
  theme(axis.title=element_text(size=14,face="bold"),
        axis.text.x=element_text(size=14,face="bold"),
        legend.title = element_text(size=14,face="bold"),
        legend.text = element_text(size=12,face="bold")) +
  stat_summary(fun.y="mean",geom="crossbar", 
               mapping=aes(ymin=..y.., ymax=..y..), width=1,show.legend = FALSE)
p2 <- ggplot(jointPC, aes(x=group, y=-U.1, color = group)) + 
  geom_sina(aes(group = group), show.legend = FALSE) +
  ylab("Glycomics (u1)")+
  theme(axis.title=element_text(size=14,face="bold"),
        axis.text.x=element_text(size=14,face="bold")) +
  stat_summary(fun.y="mean",geom="crossbar", 
               mapping=aes(ymin=..y.., ymax=..y..), width=1,show.legend = FALSE)

cairo_pdf(file = "/home/z/Dropbox/articles/Two-step/score_dot.pdf", width = 4, height =6, onefile = T, fallback_resolution = 600)
gridExtra::grid.arrange(p1,p2,nrow = 1)
dev.off()


# Dot plot on sex
ggplot(jointPC, aes(x=sex, y=Tt.1, color = sex)) + 
  geom_sina(aes(group = sex)) +
  ylab("Methylation (T1)")+
  theme(axis.title=element_text(size=14,face="bold"),
        axis.text.x=element_text(size=14,face="bold"),
        legend.title = element_text(size=14,face="bold"),
        legend.text = element_text(size=12,face="bold")) +
  stat_summary(fun.y="mean",geom="crossbar", 
               mapping=aes(ymin=..y.., ymax=..y..), width=1,show.legend = FALSE)
ggplot(jointPC, aes(x=sex, y=U.1, color = sex)) + 
  geom_sina(aes(group = sex)) +
  ylab("Glycomics (U1)")+
  theme(axis.title=element_text(size=14,face="bold"),
        axis.text.x=element_text(size=14,face="bold")) +
  stat_summary(fun.y="mean",geom="crossbar", 
               mapping=aes(ymin=..y.., ymax=..y..), width=1,show.legend = FALSE)


gridExtra::grid.arrange(p1 + theme(legend.position="none"),
                        p2 + theme(legend.position="none"),
                        nrow=1, ncol=2)
```


## Linear mixed model: Joint PC ~ DS + age + sex + (1|family)
Linear mixed model with random intercept for family gives much more significant results comparing to linear model. 
T1, T3, U1, U3 are significant
```{r}
fit_mixed_t1 <- lmerTest::lmer(-Tt.1 ~ group_ds +age + sex + age*group_ds + (1|family), data=jointPC)
fit_mixed_t2 <- lmerTest::lmer(Tt.2 ~ group_ds +age +sex + age*group_ds + (1|family), data=jointPC)
fit_mixed_t3 <- lmerTest::lmer(-Tt.3 ~ group_ds +age +sex + age*group_ds +  (1|family), data=jointPC)
fit_mixed_u1 <- lmerTest::lmer(U.1 ~ group_ds +age+ sex + age*group_ds + (1|family), data=jointPC)
fit_mixed_u3 <- lmerTest::lmer(U.3 ~ group_ds +age +sex + age*group_ds + (1|family), data=jointPC)

summary(fit_mixed_t1)
summary(fit_mixed_t2)
summary(fit_mixed_t3)
summary(fit_mixed_u1)
summary(fit_mixed_u3)
###
capture.output(list(summary(fit_mixed_t1),summary(fit_mixed_t3),summary(fit_mixed_u1),
                    summary(fit_mixed_u3)), file = "temp_out.txt")


lm(-T1_con~ group_ds +age_con + sex + age_con*group_ds, data = jointPC) %>% summary

jointPC %<>% group_by(family) %>% mutate(T1_con =Tt.1-mean(Tt.1), age_con = age-mean(age))


fa_itercpt_t1 <- ranef(fit_mixed_t1)$family[as.character(jointPC$family),]
fa_itercpt_t3 <- ranef(fit_mixed_t3)$family[as.character(jointPC$family),]
fa_itercpt_u1 <- ranef(fit_mixed_u1)$family[as.character(jointPC$family),]
fa_itercpt_u3 <- ranef(fit_mixed_u3)$family[as.character(jointPC$family),]


jointPC %<>% mutate(cor_T1 = Tt.1-fa_itercpt_t1,
                    cor_T3 = Tt.3-fa_itercpt_t3,
                    cor_U1 = U.1-fa_itercpt_u1,
                    cor_U3 = U.3-fa_itercpt_u3)


jointPC %<>% mutate(cor_all_T1 = lm(cor_T1~sex, data=jointPC)$residual,
                    cor_all_T3 = lm(cor_T3~sex, data=jointPC)$residual,
                    cor_all_U1 = lm(cor_U1~sex, data=jointPC)$residual,
                    cor_all_U3 = lm(cor_U3~sex, data=jointPC)$residual)

cor(jointPC$cor_all_T1, jointPC$cor_all_U1)
cor(jointPC$cor_T1, jointPC$cor_U1)
cor(jointPC$Tt.1, jointPC$U.1)


# diagnostics
## Residual
plot(fit_mixed)
qqnorm(residuals(fit_mixed))
## Linearity in age
ggplot(data.frame(x1=jointPC$age,pearson=residuals(fit_mixed,type="pearson")),
      aes(x=x1,y=pearson)) +
    geom_point() +
    theme_bw()

```


# loadings
```{r}
cpg_T1 <- with(fit_po2$parameters, colnames(meth)[order(W[,1]^2, decreasing = T)[1:1000]])

genes_T1 <- (probe %>% filter(ID %in% cpg_T1))$UCSC_RefGene_Name
genes_T1 <- sapply(strsplit(genes_T1, ";"), function(e) e)
genes_T1 <- do.call(c, genes_T1) %>% unique
genes_T1 <-genes_T1[!is.na(genes_T1)]

write.table(genes_T1, file = "genes_meth_T1_old.txt", quote = F, row.names = F, col.names = F)

```
